---
title: 'EDS 241: Assignment 3'
author: "Alexandra Yousefivand"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: 
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
  \setlength{\parindent}{1em}
  \usepackage{float}
  \renewcommand{\thesection}{Question (\alph{section})}
---

``` {r setup, echo = FALSE, message = FALSE, warning = FALSE}

# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3,
                      echo = TRUE, message = FALSE, warning = FALSE)                     
```


```{r}
# load packages
packages=c("readxl", "stargazer", "here", "tidyr",
           "dplyr","stringr", "janitor",
           "cowplot", "ggplot2", "tinytex",
           "datasets", "tibble", "estimatr")

for (i in packages) {
  if (require(i,character.only=TRUE)==FALSE) {
    install.packages(i,repos='http://cran.us.r-project.org')
  }
  else {
    require(i,character.only=TRUE)
  }
}

#devtools::install_github('rstudio/rmarkdown')
options(scipen=999) # not scientific notation
```


```{r load data}
# load data
data_raw <- read.csv("SMOKING_EDS241.csv")

data <- data_raw %>% clean_names()
```


# (a)

_What is the unadjusted mean difference in birth weight of infants with smoking and nonsmoking mothers?_


```{r subset data}
# subset smoking mothers
data_smoking <- data %>% filter(tobacco == 1)

# subset nonsmoking mothers
data_non_smoking <- data %>% filter(tobacco == 0)
```


```{r}
unadjusted_mean_difference <-
  mean(data_non_smoking$birthwgt) - mean(data_smoking$birthwgt)
```

**The unadjusted mean difference is `r round(unadjusted_mean_difference, 2)` grams.**


_Under what hypothesis does this correspond to the average treatment effect of maternal smoking during pregnancy on infant birth weight? Provide some simple empirical evidence for or against this hypothesis._

**Treatment ignorability**:
all other confounding variables (education, race, alcohol, first, diabete, anemia) are held constant.







# (b)


_Assume that maternal smoking is randomly assigned conditional on the observable covariates listed above. Estimate the effect of maternal smoking on birth weight using a linear regression. Report the estimated coefficient on tobacco and its standard error._


```{r tobacco coefficient, random}
model <- lm_robust(formula = birthwgt ~ tobacco,
                   data = data)
summary(model)
```

```{r}
tobacco_coefficient <- model$coefficients[[2]]
tobacco_coefficient

tobacco_std_error <- model$std.error[[2]]
tobacco_std_error
```



# (c)


_Use the exact matching estimator to estimate the effect of maternal smoking on birth weight. For simplicity, consider the following covariates in your matching estimator: create a 0-1 indicator for mother's age (=1 if mage>=34), and a 0-1 indicator for mother's education (1 if meduc>=16), mother's race (mblack), and alcohol consumption indicator (alcohol). These 4 covariates will create 2 * 2 * 2 * 2 = 16 cells. Report the estimated average treatment effect of smoking on birthweight using the exact matching estimator and its linear regression analogue (Lecture 6, slides 12-14)._



```{r}
data_indicator <- data %>%
  mutate(age_above_34 = case_when(mage >= 34 ~ 1,
                                  mage <  34 ~ 0),
        educ_above_16 = case_when(meduc >= 16 ~ 1,
                                  meduc <  16 ~ 0),
        g = as.factor(paste0(age_above_34,
                             educ_above_16,
                             mblack,
                             alcohol)))
```




```{r}
model <- lm(formula = birthwgt ~ age_above_34 + educ_above_16 + mblack + alcohol,
            data = data_indicator)
summary(model)
```


```{r exact matching estimator}
TIA <- data_indicator %>%
  group_by(g, tobacco) %>% 
  summarise(n_obs = as.integer(n()),
            birthwgt_mean = mean(birthwgt, na.rm = TRUE)) %>%
  gather(variables, values, n_obs:birthwgt_mean) %>% 
  mutate(variables = paste0(variables, "_", tobacco)) %>% 
  pivot_wider(id_cols = g, names_from = variables, values_from = values) %>%
  ungroup() %>% 
  mutate(diff = birthwgt_mean_1 - birthwgt_mean_0,
         w_ATE = (n_obs_0 + n_obs_1) / (sum(n_obs_0) + sum(n_obs_1)),
         w_ATT = n_obs_1 / sum(n_obs_1)) %>% 
  mutate_if(is.numeric, round, 2)
```

```{r}
head(TIA)
```


```{r}
# Average Treatment Effect
ATE = sum(TIA$w_ATE * TIA$diff)
ATE

# Average Treatment effect on the Treated
ATT = sum(TIA$w_ATT * TIA$diff)
ATT
```

```{r}
model <- lm_robust(formula = birthwgt ~ tobacco + g, data = data_indicator)
summary(model)
```


```{r}
tobacco_coefficient <- model$coefficients[["tobacco"]]
tobacco_coefficient
```


# (d)

_Estimate the propensity score for maternal smoking using a logit estimator and based on the following specification: mother’s age, mother’s age squared, mother’s education, and indicators for mother’s race, and alcohol consumption._


```{r}
data_logit <- data_indicator %>% mutate(mage_squared = mage^2)
```


```{r}
model_logit <- glm(formula = tobacco ~ mage +
                                       mage_squared +
                                       meduc +
                                       mblack +
                                       alcohol,
                   family = binomial(link = "logit"),
                   data = data_logit)
summary(model_logit)
```


```{r}
EPS <- predict(model_logit, type = "response")
```


```{r propensity score}
PS_WGT <- (data_logit$tobacco / EPS) + ((1 - data_logit$tobacco) / (1 - EPS))
```


```{r}
head(PS_WGT)
```




# (e)

_Use the propensity score weighted regression (WLS) to estimate the effect of maternal smoking on birth weight (Lecture 7, slide 12)._



```{r}
model_ps <- lm_robust(formula = birthwgt ~ tobacco +
                                           mage +
                                           mage_squared +
                                           meduc +
                                           mblack +
                                           alcohol,
                      weights = PS_WGT,
                      data = data_logit)
summary(model_ps)
```




```{r}
tobacco_coeff_ps <- model_ps$coefficients[["tobacco"]]
tobacco_coeff_ps
```






